A real-time particle filter for modeling travel times of transit vehicles through a GTFS-based road network

### 2018-11-01 anzjs2018/sections/01_introduction.tex
edits

In public transport, \rt information (RTI), such as estimated times of arrival (ETAs),
keeps transport users informed about their journey,
and potentially make decisions on the best route to take.
Unfortunately, here in Auckland, RTI is highly unreliable,
and succumbs to several problems that public transport users
are all too familiar with.
One of these is inaccurate ETAs,
which tend to either increase or decrease over time as the bus gets progressively later or earlier
due to traffic congestion, or  because the schedule is poorly calibrated.
Another is that buses are shown as \emph{on-time} if they are not present
in the \rt system,
regardless of whether this is due to technical problems,
or the bus failing to show up;
in either case, schedule times are used to determine ETAs and, 
if the bus doesn't show up on time,
the service is removed from the \rt board 
and we get what passengers have described as ``Ghost Buses''.

@@

To avoid the problems associated with the arrival time predictions
based solely on reported delays,
a more robus modeling and prediction framework is required.
This should consist of a robust vehicle model to estimate the position and speed
of transit vehicles from \rt GPS data;
and a way to incorporate road congestion information 
into the arrival time predictions.
Several vehicle modeling approaches were explored, 
including the Kalman filter \citep{Dailey_2001,Cathey_2003},
machine learning models \citep{Yu_2006,Chang_2010},
and a particle filter \citep{Hans_2015}.
The last of these was chosen for several reasons
which are discussed in Section~\ref{sec:pf}.

>>>>

In public transport, \rt information (RTI), such as estimated times of arrival (ETAs),
keeps transport users informed about their journey,
and potentially make decisions on the best route to take.
It has also been shown that perceived waiting time is less
when arrival time information is available \citep{TCRP_2003b}.
Unfortunately, here in Auckland, RTI is highly unreliable,
and succumbs to several problems that public transport users
are all too familiar with.
One of these is inaccurate ETAs,
which tend to either increase or decrease over time as the bus gets progressively later or earlier
due to traffic congestion, or  because the schedule is poorly calibrated.
Another is that buses are shown as \emph{on-time} if they are not present
in the \rt system,
regardless of whether this is due to technical problems,
or the bus failing to show up;
in either case, schedule times are used to determine ETAs and, 
if the bus doesn't show up on time,
the service is removed from the \rt board 
and we get what passengers have described as ``Ghost Buses''.

@@

To avoid the problems discussed above,
a more robust modeling and prediction framework is required.
This should consist of a robust vehicle model to estimate the position and speed
of transit vehicles from \rt GPS data;
and a way to incorporate road congestion information 
into the arrival time predictions.
Several vehicle modeling approaches were explored, 
including the Kalman filter \citep{Dailey_2001,Cathey_2003},
machine learning models \citep{Yu_2006,Chang_2010},
and a particle filter \citep{Hans_2015}.
The last of these was chosen for several reasons
which are discussed in Section~\ref{sec:pf}.

### 2018-11-01 anzjs2018/sections/02_gtfs.tex
edits

\begin{figure}[tb]
    \centering
    \begin{subfigure}{0.7\textwidth}
        \centering
        \includegraphics[width=0.95\textwidth]{figures/02_network_segments_1.pdf}
        \caption{Raw GTFS route shapes}
        \label{fig:network_creation_1}
    \end{subfigure} \\
    \begin{subfigure}{0.7\textwidth}
        \centering
        \includegraphics[width=0.95\textwidth]{figures/02_network_segments_2.pdf}
        \caption{GTFS-based road network}
        \label{fig:network_creation_2}
    \end{subfigure}
    \caption{Construction of a route network involves combining routes at nodes %
        (here we are using stops) which are connected by edges (roads). %
        In (a), the three unique routes, represented by different line types, clearly %
        overlap in several places. In (b) these have been merged, and the width of each line %
        represents how many routes use that link.}
    \label{fig:network_creation}
\end{figure}

>>>>

\begin{figure}[tb]
    \centering
    \begin{subfigure}{0.7\textwidth}
        \centering
        \includegraphics[width=0.95\textwidth]{figures/02_network_segments_1.pdf}
        \caption{Raw GTFS route shapes}
        \label{fig:network_creation_1}
    \end{subfigure} \\
    \begin{subfigure}{0.7\textwidth}
        \centering
        \includegraphics[width=0.95\textwidth]{figures/02_network_segments_2.pdf}
        \caption{GTFS-based road network}
        \label{fig:network_creation_2}
    \end{subfigure}
    \caption{
        Constructing a route network from the stops (dots) of three separate
        routes (lines), as shown in (a).
        Routes using the same subsequence of stops are assigned the same
        road segments, so the road network shown 
        consists of nodes (stops) and edges (roads),
        as shown in (b), where the thickness of each lines
        corresponds to how many routes share that road.
    }
    % Construction of a route network involves combining routes at nodes %
    %     (here we are using stops) which are connected by edges (roads). %
    %     In (a), the three unique routes, represented by different line types, clearly %
    %     overlap in several places. In (b) these have been merged, and the width of each line %
    %     represents how many routes use that link.}
    \label{fig:network_creation}
\end{figure}

### 2018-11-01 anzjs2018/sections/03_models.tex
edits

\begin{figure}[tb]
    \centering
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/03_particle_filter_1.pdf}
        \caption{The vehicle state represented by a sample of discrete points is related to the observable state (GPS position) through the transition function $h$.}
        \label{fig:pf_state_prev}
    \end{subfigure}\;\;
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/03_particle_filter_2.pdf}
        \caption{The transition function $f$ models the behaviour of each particle,
            including changes in acceleration and stopping behaviour at bus stops,
            which can result in multiple modes.}
        \label{fig:pf_state_predict}
    \end{subfigure}\\
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/03_particle_filter_6.pdf}
        \caption{Each particle's distance from the true observation (red cross) 
            is calculated using
            the geographic distance between two coordinates, which can then be used
            to calculate it's weight.}
        \label{fig:pf_state_update}
    \end{subfigure}\;\;
    \begin{subfigure}[t]{0.48\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/03_particle_filter_4.pdf}
        \caption{Resampling occurs using likelihood weights that use the distance
            between the particle's location and the vehicles reported location (red cross),
            resulting in the updated state.}
        \label{fig:pf_state_predict2}
    \end{subfigure}
    \caption{The vehicles state is estimated using a set of discrete points, or particles,
        each of which is independently mutated to make predictions of the future state.
        Multimodality is easily handled, as demonstrated here,
        without degeneration of the filter.}
    \label{fig:pf_state}
\end{figure}

>>>>

\begin{figure}[p]
    \centering
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_1.pdf}
        \caption{
            Vehicle state is mapped to observation space by the
            measurement function $h$.   
        }
        % The vehicle state represented by a sample of discrete points is related to the observable state (GPS position) through the transition function $h$.}
        \label{fig:pf_state_prev}
    \end{subfigure}\\
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_2.pdf}
        \caption{
            The transition function $f$ predicts the future state 
            of each particle.
        }
            % The transition function $f$ models the behaviour of each particle,
            % including changes in acceleration and stopping behaviour at bus stops,
            % which can result in multiple modes.}
        \label{fig:pf_state_predict}
    \end{subfigure}\\
    % \begin{subfigure}[t]{0.48\textwidth}
    %     \centering
    %     \includegraphics[width=\textwidth]{figures/03_particle_filter_6.pdf}
    %     \caption{Each particle's distance from the true observation (red cross) 
    %         is calculated using
    %         the geographic distance between two coordinates, which can then be used
    %         to calculate it's weight.}
    %     \label{fig:pf_state_update}
    % \end{subfigure}\;\;
    \begin{subfigure}[t]{0.9\textwidth}
        \centering
        \includegraphics[width=0.7\textwidth]{figures/03_particle_filter_4.pdf}
        \caption{
            The state is updated by weighted resampling based on distance
            of particle to observation.
        }
        % Resampling occurs using likelihood weights that use the distance
        %     between the particle's location and the vehicles reported location (red cross),
        %     resulting in the updated state.}
        \label{fig:pf_state_predict2}
    \end{subfigure}
    \caption{
        The particle filter approximates vehicle state using a set of 
        discrete points (a), which are each mutated indepently (b)
        to predict the next state.
        The measurement function $h$ allows state particle state
        to be mapped to a GPS coordinate (a),
        which can be compared to the observed location (c)
        to calculate each particles likelihood and perform resampling.
    }
        % The vehicles state is estimated using a set of discrete points, or particles,
        % each of which is independently mutated to make predictions of the future state.
        % Multimodality is easily handled, as demonstrated here,
        % without degeneration of the filter.}
    \label{fig:pf_state}
\end{figure}

\afterpage{\clearpage}

### 2018-11-01 anzjs2018/sections/04_implementation.tex
edits

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/04_model_results_timing.pdf}
    \caption{The timings for various parts of the application, and overall. %
        The trend is slightly non-linear due to the complexity of the ETA generation step, %
        which requires sorting of the particle ETAs to obtain quantile estimates.}
    \label{fig:timings}
\end{figure}

@@

For each simulation, 
the model was evaluated using several statistics.
These were effective sample size, $N_\text{eff}$ calculated using (\ref{eq:neff});
degeneration rate, the percentage of iterations in which the vehicle was lost
(no plausible particles);
and the variance of segment travel time estimates.

The GPS error parameter values were chosen by examining the distribution
of the distance between observations and the nearest point on the route.
This is shown in Figure~\ref{fig:dist_to_route},
in which we see a peak at about 1m and another at 2.5m,
so we have chosen GPS error values of $\epsilon \in \{1,2,3,5\}$.
System error is the variation of speed per second,
or acceleration rate,
so our chosen values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$ 
correspond to a variance of 0.09, 0.9, 9, and 45 meters per second over 
30~seconds.

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/04_model_results_dist.pdf}
    \caption{The distance between the observation and the nearest point on the shape.}
    \label{fig:dist_to_route}
\end{figure}

@@

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/04_model_results_degen.pdf}
    \caption{The effective sample size and degeneration rate for a range of parameter 
        values. The columns are of increasing number of particles,
        and the rows are for increasing (downwards) GPS error.}
    \label{fig:degen_rate}
\end{figure}

The last measure of performance is the variance of segment travel times.
This is calculated by \_\_\_.
Figure~\ref{fig:travel_times} shows that variance increases with GPS error,
and decreases with $N$,
with little effect made by system noise.
There is therefore a tradeoff between particle filter performance
($N_\text{eff}$ and degeneration rate) and parameter estimation,
although the true test will be calculating the prediction error
of ETAs once the prediction model has been implemented.

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/04_model_results_times.pdf}
    \caption{The within-segment posterior travel time uncertainty (standard error), subset by number of particles.}
    \label{fig:travel_times}
\end{figure}

>>>>

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/04_model_results_timing.pdf}
    \caption{
        Application timings were recorded for each component of each
        iteration, and averages over a 15~minute window are shown here.
        The timings shown are \emph{wall time}, the actual time passed,
        not CPU time which would be approximately 8~times larger 
        (for using 8~cores).
        Steps that only involve particle calculations are approximately
        of complexity $O(N)$,
        however some steps are a little more complex, 
        so there is a slightly exponential trend.
    }
        % The timings for various parts of the application, and overall. %
        % The trend is slightly non-linear due to the complexity of the ETA generation step, %
        % which requires sorting of the particle ETAs to obtain quantile estimates.}
    \label{fig:timings}
\end{figure}

@@

To evaluate the performance of the model,
the simulation was run using a range of values of GPS error, $\epsilon$,
and system noise, $\sigma^2$.
Figure \ref{fig:dist_to_route} shows the distribution of the distance
between each observation and the route using a nearest point algorithm,
so we chose to use GPS error values $\epsilon \in \{1,2,3,5\}$. 
For the system noise we use values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
which correspond to the average vehicle speed having a variance between 0.1 and 45~meters per second
over a period of 30~seconds.
For each of the combination of these parameter values, along with varying $N$,
several values were computed.
These were effective sample size, $N_\text{eff}$ calculated using (\ref{eq:neff});
degeneration rate, the percentage of iterations in which the vehicle was lost
(no plausible particles);
and the variance of segment travel time estimates.

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/04_model_results_dist.pdf}
    \caption{
        The distance between the observation and the nearest point on the shape
        is calculated by finding the shortest distance between point and path.
        This gives an approximate idea of GPS error, however it should be noted
        that many observations are made automatically at waypoints,
        which accounts for the peak at 1m.
        Note also that these observations have been truncated at 20m.
    }
    \label{fig:dist_to_route}
\end{figure}

@@

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/04_model_results_neff.pdf}
    \caption{
        The effective sample size is an approximate measure of the variance
        of the particle weights.
        When all particles have equal weights, $N_\text{eff} = N$,
        while if all the weight is on one particle, $N_\text{eff} = 1$. 
        The figure is subset by system noise, which shows how $N_\text{eff}$ 
        decreases as system noise increases,
        while increasing GPS error (on the $x$-axis) increases $N_\text{eff}$.
        Increasing $N$ has a negative effect on $N_\text{eff}$ but this is
        only evident when $N_\text{eff}$ is larger.
        The error bars show $\pm 2$SE.
    }
    \label{fig:neff}
\end{figure}

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/04_model_results_degen.pdf}
    \caption{
        The degeneration rate is the percentage of iterations where there were no
        plausible particles, so the vehicle was ``lost'' by the model.
        This decreases as GPS error and number of particles increase,
        but there is no visible effect of system noise.
        The error bars show $\pm 2$SE.
    }
    \label{fig:degen_rate}
\end{figure}

The last measure of performance is the relative variance of segment travel times.
This is calculated by computing the average variance of travel times over all simulations,
and then for each simulation, 
taking the average (over road segments) of the ratio between variance for that simulations
and variance for all simualtions.
Figure~\ref{fig:travel_times} shows that simulations with larger GPS error
result in a higher relative variance,
while simulations with more particles have lower relative variance.
There is no visible effect of system noise on travel time variation.

The results of the simulations
demonstrate a tradeoff between particle filter performance
($N_\text{eff}$ and degeneration rate) and parameter estimation.
However, until the arrival time prediction model has been completed,
it is difficult to make decisions about the optimal values to use:
we are unable to say whether a higher variation of travel time estimates
is going to have a significant effect on the final ETAs,
particularly when compared to the uncertainty of forecasting road state.

\begin{figure}[tb]
    \centering
    \includegraphics[width=\textwidth]{figures/04_model_results_times.pdf}
    \caption{
        The variability of travel time within a segment increases with GPS error,
        and decreases slightly with increasing number of particles.
        There is no visible effect of system noise on travel time variability.
        The error bars show $\pm 2$SE.
    }
    \label{fig:travel_times}
\end{figure}

### 2018-11-01 anzjs2018/sections/06_discuss.tex
edits

In this paper we have described an approach to transit vehicle modeling
that enables the \rt estimation of road state from vehicle position data,
which will in future be used to make what will hopefully be more reliable 
arrival time predictions.
The vehicle model, implemented using a particle filter,
allows complex bus behaviours to be modeled,
with future work planned to explore the other parameters---dwell time and stopping probabilities---%
as well as to extend the model to also account for physical intersections.
The travel times estimated from the particle filter then allow modeling of
\rt road state.

The network model presented here is simple,
but provides the foundation for future work which will involve
developing a formal model that combines \rt information with
historical data, allowing for more accurate predictions of future road state,
for example incorporating peak time congestion forecasts.
This in turn will provide the basis for an arrival time prediction model,
a combination of vehicle state and road state forecasts.

The primary purpose of this paper was to present our approach and 
explore and demonstrate tis faesibility as a \rt option for RTI in transit systems.
We have been able to construct an application that runs in \rt and capable
of producing estimates within 30~seconds---although this can be reduced 
by using fewer particles or, if necessary, increasing the number of cores.

>>>>

In this paper we have described an approach to transit vehicle modeling
that enables the \rt estimation of road state from vehicle position data,
which can later be used to make arrival time predictions
that account for \rt traffic conditions.
The vehicle model, implemented using a particle filter,
allows complex bus behaviours to be modeled,
with future work planned to extend the model to handle physical intersections,
such as traffic lights.
From the particle filter,
vehicle travel times along roads are estimated which are in turn used 
in the modeling of the road network state.

The network model presented here is simple,
but provides the foundation for future work to
devel a formal model combining \rt travel time information with historical data, 
allowing short- and medium-term forecasts of road state,
for example incorporating peak time congestion.
This in turn will provide the basis for an arrival time prediction model
that combines \rt vehicle state with travel time forecasts
and provide commuters with reliable ETAs.
These will likely consist of a point estimate along with a prediction interval,
as it is impossible given the available data to accurately predict arrival time
due to traffic lights, unknown dwell times at intermediate bus stops,
and many other factors.
